Do we need detailed demographic data to forecast plant population responses to climate change?

### 2014-12-21 manuscript/manuscript.Rmd
updated ms with equations

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit analagous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species. We describe the statistical models seprately since fitting the models required different approaches at the individual and quadrat levels. All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year

@@

\begin{equation}
\text{logit}(S_{ijQ,t}) = \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t}
\end{equation}

where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species. $\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat $Q$.

@@

\begin{equation}
x_{ijQ,t+1} = \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t}
\end{equation}

@@

Our data allows us to track new recruits, but we cannot assign a specific parent to new genets. So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$). Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($Q$) of quadrats in which $q$ is located:

@@

We assume the number of individuals, $y$, recruiting at time $t+1$ follows a negative binomial distribution:

\begin{equation}
y_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
\end{equation}

@@

\begin{equation}
\lambda{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
\end{equation}

@@

\begin{equation}
\text{logit}(S_{jq,t}) = \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t}
\end{equation}

@@

\begin{align}
\rho &= x_{jq,t+1}\tau_{j} \\
\eta &= (1-x_{jq,t+1})\tau_{j}
\end{align}

### Model fitting
All of our analyses (model fitting and simulating) were conducted in R [@Team2013]. We used the MCMC sampler in JAGS [@Plummer2003] to estimate the posterior distributions of model parameters and the package 'r2jags' [@Su2012] to connect R to JAGS. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding an initial 50,000 iterations. We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006]. Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

@@

Results
=======
We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only. When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability (@Adler2012).

>>>>

### Stastical models of vital rates
At both levels of inference (individual and quadrat), the building blocks of our population models are vital rate regressions. For individual level data we fit models for survival, growth, and recruitment of new individuals for each species. At the quadrat level we fit analagous models of extinction probability, percent cover increase/decrease, and quadrat colonization for each species. We describe the statistical models seprately since fitting the models required different approaches at the individual and quadrat levels. All models contain four climate covariate that we chose \emph{a priori}: fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and ppt2, respectively) and mean spring temperature at \emph{t}-1 and \emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is the observation year.

We fit all models using a hierarchical Bayesian approach, which we describe in more detail below. However, for each vital rate statistical model we also define the likelihood model we use. For the likelihood models, Y is always the relevant vector of observations (e.g., whether a  genet survived [1] or not [0] from year $t$ to $t+1$).

@@

\begin{align}
\text{logit}(S_{ijQ,t}) &= \gamma^{S}_{j,t} + \phi^{S}_{jQ} + \beta^{S}_{j,t}x_{ij,t} + \omega^{S}_{j}w_{ij,t} + \theta^{S}_{jk}C_{k,t} + \varepsilon^{S}_{t} \\
Y^{S}_{ijQ,t} &\sim \text{Bernoulli}(S_{ijQ,t})
\end{align}

where $x_{ij,t}$ is the log of genet size, $\gamma^{S}_{j,t}$ is a year-specific intercept, $\beta^{S}_{j,t}$ is the year-specific slope parameter for size, $\phi^{S}_{jQ}$ is the random effect of quadrat group location, and $\theta^{S}_{k}$ is the fixed parameter for the effect of the $k$th climate covariate at time $t$ ($C_{k,t}$). We include density-dependence by estimating the effect of crowding on the focal individual by other individuals of the same species. $\omega$ is the effect of crowding and $w_{t,Q}$ is the crowding experienced by the focal individual at time $t$ in quadrat group $Q$.

@@

\begin{align}
x_{ijQ,t+1} &= \gamma^{G}_{j,t} + \phi^{G}_{jQ} + \beta^{G}_{j,t}x_{ij,t} + \omega^{G}_{j}w_{ij,t} + \theta^{G}_{jk}C_{k,t} \\
Y^{G}_{ijQ,t} &\sim \text{Normal}(x_{ijQ,t+1}, \sigma_{j})
\end{align}

@@

Our data allows us to track new recruits, but we cannot assign a specific parent to new genets. So, for recruitment, we work at the quadrat level and model the number of new individuals of species $j$ in quadrat $q$ recruiting at time $t+1$ as a function of quadrat "effective cover" ($A'$) in the previous year ($t$). Effective cover is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which $q$ is located:

@@

We assume the number of individuals, $Y^{R}$, recruiting at time $t+1$ follows a negative binomial distribution:

\begin{equation}
Y^{R}_{jq,t+1} \sim \text{NegBin}(\lambda_{jq,t+1},\zeta)
\end{equation}

@@

\begin{equation}
\lambda_{jq,t+1} = A'_{jq,t}e^{(\gamma^{R}_{j,t} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} + \omega^{R}\sqrt{A'_{q,t}})}
\end{equation}

@@

\begin{align}
\text{logit}(S_{jq,t}) &= \gamma^{S}_{j} + \phi^{S}_{jQ} + \beta^{S}_{j}x_{jq,t} + \theta^{S}_{jk}C_{k,t} \\
Y^{S}_{jq,t} &\sim \text{Bernoulli}(S_{jq,t})
\end{align}

@@

\begin{align}
\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
\end{align}

with likelihood:

\begin{equation}
Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
\end{equation}

Finally, we modeled probability of colonization in quadrat $q$ by species $j$ from time $t$ to $t+1$ as:

\begin{align}
\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
\end{align}

### Model fitting
Our Bayesian approach to fitting the vital rate models required choosing appropriate priors for unknown parameters and deciding which, if any, of those prior should be hierarchical. We decided to fit models where all terms except climate covariates were fit by species, while the climate covariates were fit heierarchically where species-specific coefficients were drawn from a shared `global' coefficient distribution. We did so for two reasons: (1) the four focal species are all perrennial grasses that we expect to respond similarly to climate covariates, and (2) convergence of climate effects at the quadrat level was much easier to achieve when we modeled these terms hierarchically, allowing them to "share" statistical strenght via partial pooling [@Gelman2007]. So, climate effects were modeled as:

\begin{align}
\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
\end{align}

We used uninformative priors for all unknown parameters, specifically:

\begin{align}
\boldsymbol{\gamma, \beta} &\sim \text{Normal}(0, 1e-6) \\
\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
\boldsymbol{\sigma_{\phi}} &\sim e(\text{Gamma}(2, 0.5)) \\
\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}} &\sim \text{Gamma}(0.001, 0.001)
\end{align}

All of our analyses (model fitting and simulating) were conducted in R [@Team2013]. We used the MCMC sampler in JAGS [@Plummer2003] to estimate the posterior distributions of model parameters and the package 'r2jags' [@Su2012] to connect R to JAGS. We obtained posterior distributions for all model parameters from three parallel MCMC chains run for 50,000 iterations, after discarding an initial 50,000 iterations. We assessed convergence visually and using the @Gelman1992 diagnostic in the R package `coda' [@Plummer2006]. Scale reduction factors for all parameters were less than 1.02, indicating convergence. For the purposes of introducing stochasticity in our population models, we saved the final 1,000 iterations from each chain for all parameters to be used as randomly drawn values during population simulation.

@@

Results
=======
We assessed the statistical importance of the climate covariates included the final vital rate regressions by comparing the residual deviance of models with climate covariates and temporal random effects, climate covariates only, and temporal random effects only. When a model includes climate covariates, this comparison shows the relative contribution of the climate covariates in explaining the total interannual variability [@Adler2012].

### 2014-12-21 manuscript/manuscript.tex
updated ms with equations

At both levels of inference (individual and quadrat), the building
blocks of our population models are vital rate regressions. For
individual level data we fit models for survival, growth, and
recruitment of new individuals for each species. At the quadrat level we
fit analagous models of extinction probability, percent cover
increase/decrease, and quadrat colonization for each species. We
describe the statistical models seprately since fitting the models
required different approaches at the individual and quadrat levels. All
models contain four climate covariate that we chose \emph{a priori}:
fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and mean spring temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year

@@

Our data allows us to track new recruits, but we cannot assign a
specific parent to new genets. So, for recruitment, we work at the
quadrat level and model the number of new individuals of species $j$ in
quadrat $q$ recruiting at time $t+1$ as a function of quadrat
``effective cover'' ($A'$) in the previous year ($t$). Effective cover
is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
mean cover across the entire group ($Q$) of quadrats in which $q$ is
located:

@@

\begin{align}
\rho &= x_{jq,t+1}\tau_{j} \\
\eta &= (1-x_{jq,t+1})\tau_{j}
\end{align}

@@

@@

We assessed the statistical importance of the climate covariates
included the final vital rate regressions by comparing the residual
deviance of models with climate covariates and temporal random effects,
climate covariates only, and temporal random effects only. When a model
includes climate covariates, this comparison shows the relative
contribution of the climate covariates in explaining the total
interannual variability (Adler et al. (2012)).

@@

>>>>

At both levels of inference (individual and quadrat), the building
blocks of our population models are vital rate regressions. For
individual level data we fit models for survival, growth, and
recruitment of new individuals for each species. At the quadrat level we
fit analagous models of extinction probability, percent cover
increase/decrease, and quadrat colonization for each species. We
describe the statistical models seprately since fitting the models
required different approaches at the individual and quadrat levels. All
models contain four climate covariate that we chose \emph{a priori}:
fall through spring precipitation at \emph{t}-1 and \emph{t}-2 (ppt1 and
ppt2, respectively) and mean spring temperature at \emph{t}-1 and
\emph{t}-2 (TmeanSpr1 and TmeanSpr2, respectively), where \emph{t} is
the observation year.

We fit all models using a hierarchical Bayesian approach, which we
describe in more detail below. However, for each vital rate statistical
model we also define the likelihood model we use. For the likelihood
models, Y is always the relevant vector of observations (e.g., whether a
genet survived {[}1{]} or not {[}0{]} from year $t$ to $t+1$).

@@

Our data allows us to track new recruits, but we cannot assign a
specific parent to new genets. So, for recruitment, we work at the
quadrat level and model the number of new individuals of species $j$ in
quadrat $q$ recruiting at time $t+1$ as a function of quadrat
``effective cover'' ($A'$) in the previous year ($t$). Effective cover
is a mixture of observed cover ($A$) in the focal quadrat ($q$) and the
mean cover across the entire group ($\bar{A}$) of $Q$ quadrats in which
$q$ is located:

@@

\begin{align}
\rho_{jq,t+1} &= x_{jq,t+1}\tau_{j} \\
\eta_{jq,t+1} &= (1-x_{jq,t+1})\tau_{j}
\end{align}

with likelihood:

\begin{equation}
Y^{G}_{jq,t+1} \sim \text{Beta}(\rho_{jq,t+1}, \eta_{jq,t+1}).
\end{equation}

Finally, we modeled probability of colonization in quadrat $q$ by
species $j$ from time $t$ to $t+1$ as:

\begin{align}
\text{logit}(R_{jq,t}) &= \gamma^{R}_{j} + \phi^{R}_{jQ} + \theta^{R}_{jk}C_{k,t} \\
Y^{R}_{jq,t} &\sim \text{Bernoulli}(R_{jq,t}).
\end{align}

@@

Our Bayesian approach to fitting the vital rate models required choosing
appropriate priors for unknown parameters and deciding which, if any, of
those prior should be hierarchical. We decided to fit models where all
terms except climate covariates were fit by species, while the climate
covariates were fit heierarchically where species-specific coefficients
were drawn from a shared `global' coefficient distribution. We did so
for two reasons: (1) the four focal species are all perrennial grasses
that we expect to respond similarly to climate covariates, and (2)
convergence of climate effects at the quadrat level was much easier to
achieve when we modeled these terms hierarchically, allowing them to
``share'' statistical strenght via partial pooling (Gelman and Hill
2007). So, climate effects were modeled as:

\begin{align}
\theta_{jk} &\sim \text{Normal}(\bar{\theta_{k}}, \sigma_{k}) \\
\bar{\theta_{k}} &\sim \text{Normal}(0, 1e-6)
\end{align}

We used uninformative priors for all unknown parameters, specifically:

\begin{align}
\boldsymbol{\gamma, \beta} &\sim \text{Normal}(0, 1e-6) \\
\boldsymbol{\phi} &\sim \text{Normal}(0, \boldsymbol{\sigma_{\phi}}) \\
\boldsymbol{\sigma_{\phi}} &\sim e(\text{Gamma}(2, 0.5)) \\
\boldsymbol{\sigma_{\theta}, \sigma_{\gamma}, \sigma_{\beta}} &\sim \text{Gamma}(0.001, 0.001)
\end{align}

@@

We assessed the statistical importance of the climate covariates
included the final vital rate regressions by comparing the residual
deviance of models with climate covariates and temporal random effects,
climate covariates only, and temporal random effects only. When a model
includes climate covariates, this comparison shows the relative
contribution of the climate covariates in explaining the total
interannual variability (Adler et al. 2012).

@@

Gelman, A., and J. Hill. 2007. Data analysis using regression and
multilevel/hierarchical models. Page xxii, 625p.
